---
title: "graphsim: importing pathways from databases"
author: "S. Thomas Kelly^1,2^, Michael A. Black^1^ <br> ^1^ Department of Biochemistry, University of Otago, PO Box 56, Dunedin 9054, New Zealand <br> ^2^ RIKEN Center for Integrative Medical Sciences, Suehiro-cho-1-7-22, Tsurumi Ward, Yokohama"
date: "`r  format(Sys.time(), '%A %d %B %Y')`"
output:
  prettydoc::html_pretty:
       theme: cayman
  #html_document:
       #theme: united
       number_sections: true
       toc: true
       toc_depth: 4
       #toc_float: true
       #code_folding: show
       keep_html: true
vignette: >
  %\VignetteIndexEntry{Importing pathways from databases}
  %\VignetteEngine{R.rsp::asis}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", width = 68)
knitr::opts_chunk$set(fig.cap = "", fig.path = "Plot")
knitr::opts_chunk$set(fig.align = "center")
options(width = 68, cli.unicode = FALSE, cli.width = 68)
#par(mai=c(2.82, 2.82, 0.82, 0.82)-0.82)
par(mar=c(7, 10, 4, 2) + 0.1)
par(bty="o")
#captioner::captioner(prefix = "Fig.")
```


<!-- <h1 class="title toc-ignore">Using the graphsim package:  Directed plots for igraph objects</h1> -->

<!-- <h4 class="author">S. Thomas Kelly^1,2^, Michael A. Black^2^</h4> **** -->

<!-- ^1^ Department of Biochemistry, University of Otago, PO Box 56, Dunedin 9054, New Zealand -->

<!-- ^2^ RIKEN Center for Integrative Medical Sciences, Suehiro-cho-1-7-22, Tsurumi Ward, Yokohama -->

<!-- <h4 class="date">`r  format(Sys.time(), '%A %d %B %Y')`</h4> -->

<!-- ---------------------------------------------------------- -->

**Summary**

Guide on how to create `igraph` objects from pathways imported from Pathway Commons.

**Package**

graphsim 1.0.2

# Importing Pathways

## Motivations

Here we demonstrate how to create `igraph` objects <span class="citation">(Csardi and Nepusz 2006)</span>  for pathways compatible with `graphsim`. We provide example objects with the package and these examples contain additional details showing how these are imported into R. This uses the `paxtoolsr` package class="citation">(Luna et al 2015)</span> from Bioconductor.

Graph object have edge properties <span class="citation">(Barabási and Oltvai 2004)</span>. Here we show how to define the "state" parameter which can be used to differentiate inhibitions. We use different arrowheadsto show these as per convention in molecular biology. 

## Getting started

The bioconductor package to import data can be installed as follows.

```{r, include=FALSE, eval = TRUE, message=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("paxtoolsr", update = FALSE)
```

```{r, eval = FALSE, message=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("paxtoolsr")
```

To generate perform these step the following packages must be imported.

```{r, message = FALSE}
library("igraph")
library("graphsim")
library("paxtoolsr")
```


## Importing data

We will demonstrate downloading <a href="https://reactome.org/>Reactome</a> pathway <span class="citation">(Croft _et al_. 2014)</span> from the <a href="http://www.pathwaycommons.org/>Pathway Commons</a>. Reactome pathways are also available in the href="https://bioconductor.org/packages/release/data/annotation/html/reactome.db.html>reactome.db</a> package <span class="citation">(Fabregat _et al_. 2017; Ligtenberg 2019)</span> but this only contains the gene set information. We use Pathway Commons as it contains direction graph structure information for the edges.

We downlaad the Pathway Commons release into Extended Simple Interaction Format (SIF) Network format. We use a legacy version (y) to reproduce the results.

```{r, eval = TRUE}
# Importing data
results <- downloadPc2(version = 7, selectedFileName = "Pathway%20Commons.7.Reactome.BIOPAX.owl.gz")
```

However we recommend using the latest version which is:

```{r, include=FALSE}
latest_version <- sort(as.numeric(sapply(strsplit(grep("<a href=\"v", readLines("http://www.pathwaycommons.org/archives/PC2/"), value = TRUE), "v"), function(x)  strsplit(x[3], "/")[[1]][1])), decreasing = TRUE)[1]
```
```{r}
print(latest_version)
```

Run the following code to download the new version.

```{r, include = FALSE}
# Importing data
print(paste0("results <- downloadPc2(version = ", latest_version,", selectedFileName = \"PathwayCommons", latest_version, ".reactome.BIOPAX.owl.gz\")"))
```

The results will be cached here:

```{r, eval = FALSE, include = TRUE}
Sys.getenv("PAXTOOLSR_CACHE")
```
```{r, eval = TRUE, include = FALSE}
print(paste0("$HOME/.paxtoolsRCache"))
```

### Searching results

We then query the results to find the pathways of interest. See the paxtoolsr vignette for details.

```{r, warning = FALSE, message = FALSE}
## Search Pathway Commons for 'PI3K'-related pathways
searchResults <- searchPc(q = "PI3K", type = "pathway", verbose = TRUE)
pathways <- xpathSApply(searchResults, "/searchResponse/searchHit/name", xmlValue)
length(pathways)
head(pathways)
```

### Downloading a Pathway

We can create a local OWL file in this manner.


```{r, warning = FALSE, message = FALSE}
## Search Pathway Commons for 'PI3K'-related pathways
searchResults <- searchPc(q = "PI3K Cascade", type = "pathway", verbose = TRUE)
pathways <- xpathSApply(searchResults, "/searchResponse/searchHit/name", xmlValue)
length(pathways)
head(pathways)
```

We then select the first pathway.

```{r}
pathway <- pathways[1]
pathway
```

We save it to a temporary OWL file. First we extract the required columns.

```{r}
library("plyr")

#convert to data frame
searchResultsDf <- ldply(xmlToList(searchResults), data.frame)
dim(searchResultsDf)
# Simplified results
simplifiedSearchResultsDf <- searchResultsDf[, c("name", "uri", "biopaxClass")]
head(simplifiedSearchResultsDf)
```

Then we write to a temp file.

```{r}
## Use an XPath expression to extract the results of interest. In this case, the
## URIs (IDs) for the pathways from the results
tmpSearchResults <- xpathSApply(searchResults, "/searchResponse/searchHit/uri", xmlValue)

## Generate temporary file to save content into
biopaxFile <- "bioxpax-reactome-pi3k-cascade.owl"

## Extract a URI for a pathway in the search results and save into a file
idx <- which(grepl("reactome", simplifiedSearchResultsDf$uri) & grepl("PI3K Cascade", 
    simplifiedSearchResultsDf$name, ignore.case = TRUE))
uri <- simplifiedSearchResultsDf$uri[idx]
saveXML(getPc(uri, format = "BIOPAX"), file = biopaxFile)
```

### Create SIF object

We convert to th eExtended Simple Interaction Format (SIF) Network format. This gives a matrix of nodes for genes and edges for relationships bewteen.

```{r}
resultsSIF <- toSifnx(inputFile = biopaxFile)
print(paste(c("nodes:", nrow(resultsSIF$nodes))))
print(paste(c("edges:", nrow(resultsSIF$edges))))
```

With node properties:

```{r}
results.nodesDF <- as.data.frame(resultsSIF$nodes) 
head(results.nodesDF)
```

With edge properties:

```{r}
results.edgesDF <- as.data.frame(resultsSIF$edges) 
head(results.edgesDF)
```

We see that all genes and edges belong to the same Reactome pathway.

```{r}
table(results.edgesDF$PATHWAY_NAMES)
table(results.edgesDF$INTERACTION_DATA_SOURCE)
```

Edges are defined in several ways (some directional).

```{r}
table(results.edgesDF$INTERACTION_TYPE)
```
### Filtering genes and metabolites

We can then optionally filter out edges that are not related to genes or proteins.

Either by filtering edges involving metabolites.

```{r}
results.edgesDF <- results.edgesDF[results.edgesDF$INTERACTION_TYPE != "chemical-affects",]
results.edgesDF <- results.edgesDF[results.edgesDF$INTERACTION_TYPE != "reacts-with",]
results.edgesDF <- results.edgesDF[results.edgesDF$INTERACTION_TYPE != "used-to-produce",]
table(results.edgesDF$INTERACTION_TYPE)
```

Ions can be removed as follows while retaining other metabolites.

```{R}
results.edgesDF <- results.edgesDF[results.edgesDF[,1] != "2+",1:3]
results.edgesDF <- results.edgesDF[results.edgesDF[,3] != "2+",1:3]
results.edgesDF <- results.edgesDF[results.edgesDF[,1] != "3+",1:3]
results.edgesDF <- results.edgesDF[results.edgesDF[,3] != "3+",1:3]
```

Alternatively by screening the nodes for proteins.

```{r}
table(results.nodesDF$PARTICIPANT_TYPE)
```
```{r}
#extract protein nodes
results.nodesDF <- results.nodesDF[results.nodesDF$PARTICIPANT_TYPE == "ProteinReference",]
#match to edges
results.edgesDF <- results.edgesDF[results.edgesDF$PARTICIPANT_A %in% results.nodesDF$PARTICIPANT,]
results.edgesDF <- results.edgesDF[results.edgesDF$PARTICIPANT_B %in% results.nodesDF$PARTICIPANT,]
print(paste(c("nodes:", nrow(results.nodesDF))))
print(paste(c("edges:", nrow(results.edgesDF))))
```

### Creating an igraph object

Then we create and edge list from the SIF object. First we match names between edges and participants to gene symbols.

```{r}
#extract names
gene_names <- resultsSIF$nodes$PARTICIPANT_NAME
#replace with gene symbol (if defined)
gene_names[grep("hgnc symbol:", resultsSIF$nodes$RELATIONSHIP_XREF)] <- sapply(strsplit(grep("hgnc symbol:", resultsSIF$nodes$RELATIONSHIP_XREF, value = TRUE), ":"), function(x) x[2])
gene_names
```

Match gene symbols to edge participants

```{r}
results.edgesDF$PARTICIPANT_A <- gene_names[match(results.edgesDF$PARTICIPANT_A, results.nodesDF$PARTICIPANT)]
results.edgesDF$PARTICIPANT_B <- gene_names[match(results.edgesDF$PARTICIPANT_B, results.nodesDF$PARTICIPANT)]
```

```{r}
head(results.edgesDF[,c(1, 3)])
```

Create edge list with SIF edges.

```{r}
library("igraph")
g <- graph_from_edgelist(as.matrix(results.edgesDF[,c(1, 3)]))
g
```

```{r}
library("graphsim")
plot_directed(g, arrow_clip = 0.25, col.arrow = "grey75", cex.arrow = 0.5, fill.node = "lightblue", cex.node = 1.25)
```


----------------------------------------------------------

# Session info
Here is the output of `sessionInfo()` on the system on which this document was compiled running pandoc 2.1:

```{r, echo = FALSE}
sessionInfo()
```


----------------------------------------------------------

# References

<p>Barabási, A. L., and Oltvai,  Z. N. 2004. “Network Biology: Understanding the Cell’s Functional Organization.” <em>Nat Rev Genet</em> 5 (2): 101–13.</p>

<p>Croft, D., Mundo, A. F., Haw, R.,  Milacic, M., Weiser, J., Wu, G., Caudy, M., et al. 2014. “The Reactome pathway knowledgebase.” Journal Article. <em>Nucleic Acids Res</em> 42 (database issue): D472–D477. <a href="https://doi.org/10.1093/nar/gkt1102">https://doi.org/10.1093/nar/gkt1102</a>.</p>

<p>Csardi, G., and Nepusz, T. 2006. “The Igraph Software Package for Complex Network Research.” <em>InterJournal</em> Complex Systems: 1695. <a href="http://igraph.org">http://igraph.org</a>.</p>

<p>Fabregat, A., Sidiropoulos, K., Viteri, G. et al. 2017. “Reactome pathway analysis: a high-performance in-memory approach.” <em>BMC Bioinformatics</em> 18: 1695. <a href="https://doi.org/10.1186/s12859-017-1559-2">https://doi.org/10.1186/s12859-017-1559-2</a>.</p>

<p>Ligtenberg W. 2019. “reactome.db: A set of annotation maps for reactome.” R package version 1.68.0. <a href="https://bioconductor.org/packages/release/data/annotation/html/reactome.db.html>https://bioconductor.org/packages/release/data/annotation/html/reactome.db.html</a>.</p>

<p>Luna, A., Babur, Ö., Aksoy, A. B, Demir, E., Sander, C. 2016. “PaxtoolsR: Pathway Analysis in R Using Pathway Commons.” <em>Bioinformaticsl</em> 32 (8): 1262-4. <a href="https://doi.org/10.1093/bioinformatics/btv733>https://doi.org/10.1093/bioinformatics/btv733</a>.</p>