---
title: "Directed Plots in iGraph"
author: "Tom Kelly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Directed Plots in iGraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Install

```{r, message=F}
library("igraph")
library("devtools")
devtools::install_github("TomKellyGenetics/plot.igraph")
library("plot.igraph")
```

##Set up simulated graph

```{r}
graph_edges <- rbind(c("A", "C"), c("B", "C"), c("C", "D"), c("D", "E"), c("D", "F"), c("F", "G"), c("F", "I"), c("H", "I"))
graph <- graph.edgelist(graph_edges, directed = T)
```

##Plotting

`plot_directed` with default settings uses the `layout.fruchterman.reingold` as does the default `igraph::plot.igraph`.

```{r}
plot(graph)
plot_directed(graph)
```

`plot_directed` supports customised layouts and colours:

```{r}
plot_directed(graph, layout = layout.kamada.kawai)
plot_directed(graph, fill.node = "lightblue", border.node = "royalblue")
```

Colours may also be entered as a vector for each node in `V(graph)`:

```{r}
names(V(graph))
colour_vector <- ifelse(names(V(graph)) %in% c("A", "D", "I"), 1, 2)
plot_directed(graph, fill.node = c("lightblue", "grey")[colour_vector], border.node = c("royalblue", "black")[colour_vector])
```

##Arrow customisation

The `state` parameter controls whether the links are "activating" or "inhibiting". These may be applied globally as a character or numeric:

###Activating links
* "activating"
* 1
* 0

###Inhibiting links
* "inhibiting"
* -1
* 2

```{r}
plot_directed(graph, state = "activating")
plot_directed(graph, state = "inhibiting")
plot_directed(graph, state = 1)
plot_directed(graph, state = -1)
plot_directed(graph, state = 0)
plot_directed(graph, state = 2)
```

#Vectorisation

The state parameter may also apply as a vector to each edge in E(graph) respectively.

```{r}
E(graph)
plot_directed(graph, state = c(1, 1, -1, -1, 1, -1, 1, -1))
```
