---
title: "Directed plots for igraph objects"
author: "S. Thomas Kelly^1,2^, Michael A. Black^2^ <br> ^1^ Department of Biochemistry, University of Otago, PO Box 56, Dunedin 9054, New Zealand <br> ^2^ RIKEN Center for Integrative Medical Sciences, Suehiro-cho-1-7-22, Tsurumi Ward, Yokohama"
date: "`r  format(Sys.time(), '%A %d %B %Y')`"
output:
  prettydoc::html_pretty:
       theme: cayman
  #html_document:
       #theme: united
       number_sections: true
       toc: true
       toc_depth: 4
       #toc_float: true
       #code_folding: show
       keep_html: true
vignette: >
  %\VignetteIndexEntry{Directed plots for igraph objects}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", width = 68)
knitr::opts_chunk$set(fig.cap = "", fig.path = "Plot")
knitr::opts_chunk$set(fig.align = "center")
options(width = 68, cli.unicode = FALSE, cli.width = 68)
#par(mai=c(2.82, 2.82, 0.82, 0.82)-0.82)
par(mar=c(7, 10, 4, 2) + 0.1)
par(bty="o")
#captioner::captioner(prefix = "Fig.")
```


<!-- <h1 class="title toc-ignore">Using the graphsim package:  Directed plots for igraph objects</h1> -->

<!-- <h4 class="author">S. Thomas Kelly^1,2^, Michael A. Black^2^</h4> **** -->

<!-- ^1^ Department of Biochemistry, University of Otago, PO Box 56, Dunedin 9054, New Zealand -->

<!-- ^2^ RIKEN Center for Integrative Medical Sciences, Suehiro-cho-1-7-22, Tsurumi Ward, Yokohama -->

<!-- <h4 class="date">`r  format(Sys.time(), '%A %d %B %Y')`</h4> -->

<!-- ---------------------------------------------------------- -->

**Abstract**

Guide on how to plot directed graphs from `igraph` objects to display activating and repressive states.

**Package**

graphsim 0.1.2

# Directed graph plots

## Motivations

Here we demonstrate the plotting functions that come built-in with `graphsim` and compare them them to the `plot.igraph` method provided by the `igraph` package. With all due respect to the efforts of the `igraph` developers, their plotting functions leave some things to be desired in terms of aesthetics. Here we provide additional functionality to plot directed graphs. This draws upon many functions provided by `igraph` including layout settings.

In particular, graph and network represenations in biology often require displaying edge properties. Here we have the "state" parameter which can be used to differentiate these and plot activating and inhibiting or repressing relationships differently. We use different arrowheads as per convention in biochemistry and molecular genetics. 

## Getting started

To display these plots the following packages must be imported.

```{r, message=F}
library("igraph")
library("graphsim")
```

# Toy example

Here we demonstrate the plot functions on a small toy graph.

## Set up simulated graph

First we generate a graph object using the `igraph` package.

```{r}
graph_edges <- rbind(c("A", "C"), c("B", "C"), c("C", "D"), c("D", "E"), c("D", "F"), c("F", "G"), c("F", "I"), c("H", "I"))
graph <- graph.edgelist(graph_edges, directed = T)
```

## Plotting

Then we compare plotting functions to the built-in `plot.igraph` S3 method provided for igraph objects by the `igraph` package.

`plot_directed` with default settings uses the `layout.fruchterman.reingold` as does the default `igraph::graphsim`.

```{r,  fig.align='center', out.width="80%", fig.height = 10, fig.width = 10}
plot(graph)
```
```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
plot_directed(graph)
```

Here you can see that the plotting functions display graphs in a similar layout with different aesthetic parameters. We suggest that you choose the function that suits your needs and datasets.

### Custom aesthetics

We support various aesthetic parameters to control the colour and relative size of nodes and edges.

`plot_directed` supports customised layouts and colours:

```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
plot_directed(graph, layout = layout.kamada.kawai)
plot_directed(graph, fill.node = "lightblue", border.node = "royalblue")
```

#### Vectorisation

Colours may also be entered as a vector for each node in `V(graph)`:

```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
names(V(graph))
colour_vector <- ifelse(names(V(graph)) %in% c("A", "D", "I"), 1, 2)
plot_directed(graph, fill.node = c("lightblue", "grey")[colour_vector], border.node = c("royalblue", "black")[colour_vector])
```

This functionality allow highlighting particular groups based on known properties of the graph. For examples `V(graph)$type` for bipartite graphs or partitions from Louvain (`igraph::cluster_louvain`) or Leiden (`leiden::leiden`) clustering algorithms.

### Arrow customisation

The `state` parameter controls whether the links are "activating" or "inhibiting". These can denote activation and repression, positive and negative regulation, or kinase and phosphatase activity. These may be applied globally as a character or numeric:

Activating links are displated with any of the following.

- "activating"
- `1`
- `0`

Inhibiting links are displated with any of the following.
- "inhibiting"
- `-1`
- `2`

```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
plot_directed(graph, state = "activating")
plot_directed(graph, state = "inhibiting")
plot_directed(graph, state = 1)
plot_directed(graph, state = -1)
plot_directed(graph, state = 0)
plot_directed(graph, state = 2)
```

#### Vectorisation

The state parameter may also apply as a vector to each edge in `E(graph)` respectively.

```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
E(graph)
plot_directed(graph, state = c(1, 1, -1, -1, 1, -1, 1, -1))
```

Note that by default, inhibiting relationships are highlighted with different `col.arrow` which can be controlled by the input parameter.

```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
edge_properties <- c(1, 1, -1, -1, 1, -1, 1, -1)/2 + 1.5
plot_directed(graph, state = edge_properties, col.arrow = c("#00BE67", "#F8766D")[edge_properties])
```

```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
edge_properties <- c(1, 1, -1, -1, 1, -1, 1, -1)/2 + 1.5
plot_directed(graph, state = edge_properties, col.arrow = c("#F8766D", "#CD9600", "#7CAE00", "#00BE67", "#00BFC4", "#00A9FF", "#C77CFF", "#FF61CC"), fill.node = c("#F8766D", "#CD9600", "#7CAE00", "#00BE67", "#00BFC4", "#00A9FF", "#C77CFF", "#FF61CC", "#F8766D"))
```

# Empirical examples

Here we demonstrate the plotting package to display real biological pathways from the "Reactome" database. We can import these from the `data` directory included with this package. These graphs are given for examples and convenience. Any empirical data that consists of a list of directed edges can be imported as an igraph object and handled similarly. Below are some demonstrations.

## RAF/MAP kinase cascade

Here we plot the RAF/MAP kinase cascade pathway.


```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5, warning=FALSE, message = FALSE}
graph <- identity(RAF_MAP_graph)
plot_directed(graph, col.arrow = "#00A9FF", fill.node = "springgreen1")
```

## Pi3K cascade

Here we plot the phosphoinositide-3-kinase (Pi3K) cascade pathway.

```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
graph <- identity(Pi3K_graph)
plot_directed(graph, col.arrow = "#00A9FF", fill.node = "springgreen1")
```

### Pi3k/AKT pathway

Here we plot the phosphoinositide-3-kinase activation of Protein kinase B (PKB) cascade (also known as Pi3k/AKT) pathway. States are imported as edge attributes from the imported graph.


```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5, warning=FALSE, message = FALSE}
graph <- identity(Pi3K_AKT_graph)
edge_properties <- E(graph)$state
plot_directed(graph, state = edge_properties, col.arrow = c("#00A9FF", "#F8766D")[edge_properties], fill.node = c("springgreen1"))
```

## TGFβ-Smad pathway

Here we plot the TGFβ-Smad pathway with inhibitions known. States are imported as edge attributes from the imported graph.


```{r,  fig.align='center', out.width="80%", fig.height = 5, fig.width = 5}
graph <- identity(TGFBeta_Smad_graph)
edge_properties <- E(graph)$state
plot_directed(graph, state = edge_properties, col.arrow = c("#00A9FF", "#F8766D")[edge_properties], fill.node = c("springgreen1"))
```


----------------------------------------------------------

# Session info
Here is the output of `sessionInfo()` on the system on which this document was compiled running pandoc 2.1:

```{r, echo = FALSE}
sessionInfo()
```